---
title: "Differential Expression Analysis of MCT Inhibitor RNA-seq Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE,
  eval = TRUE
)
```

## Overview

This vignette demonstrates a differential expression analysis of the RNA-seq
data in the `rnaseq.hs.lf.tgfb.mcti` package. The package contains two
experiments examining the effects of lactate transporter inhibition on
TGFβ-stimulated human lung fibroblasts:

- **`mcti`**: MCT1 inhibitor AZD3965, MCT4 inhibitor VB124, and dual
inhibition (4 replicates × 2 conditions × 4 treatments = 32 samples)
- **`vb253`**: MCT4 inhibitor VB253 (4 replicates × 2 conditions ×
2 treatments = 16 samples)

The analysis covers principal component analysis (PCA), differential expression
with DESeq2, gene set enrichment analysis (GSEA) with Hallmark pathways, and
transcription factor enrichment analysis (TFEA) with DoRothEA regulons.

```{r packages}
library(rnaseq.hs.lf.tgfb.mcti)
library(SummarizedExperiment)
library(DESeq2)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
```

## Preparing the DESeq2 datasets

Both experiments use a combined condition–treatment group factor for modeling.
The MCTi experiment includes a replicate term to account for biological batch
effects across the four independent replicates. The `/` in `AZD/VB` is replaced
with `.` to avoid conflicts with formula parsing in downstream tools.

```{r prepare-se}
prepare_dds <- function(se, design) {
  se$group <-
    str_c(se$condition, se$treatment, sep = ".") |>
    str_replace_all("/", ".") |>
    factor(levels = paste(
      rep(levels(se$condition), each = nlevels(se$treatment)),
      rep(levels(se$treatment), nlevels(se$condition)),
      sep = "."
    ) |> str_replace_all("/", "."))
  
  dds <-
    DESeqDataSet(se, design = design) |>
    DESeq()
  
  rownames(dds) <- str_replace(rownames(dds), "\\.\\d+$", "")
  
  keep <- rowSums(counts(dds, normalized = TRUE) >= 0.5) >= 4
  dds[keep, ]
}
```

```{r fit-deseq2}
data(mcti)
data(vb253)

dds_mcti <- prepare_dds(mcti, ~ replicate + group)
dds_vb253 <- prepare_dds(vb253, ~ group)
```

## Principal component analysis

Variance-stabilized counts are used for PCA. For the MCTi experiment, replicate
batch effects are removed with `limma::removeBatchEffect()` prior to
visualization.

```{r pca-function}
plot_pca <- function(dds, remove_batch = FALSE) {
  vsd <- vst(dds, blind = FALSE)
  
  if (remove_batch) {
    assay(vsd) <- limma::removeBatchEffect(assay(vsd), vsd$replicate)
  }
  
  pca_data <- plotPCA(vsd, intgroup = c("condition", "treatment"),
                      returnData = TRUE)
  pct <- round(100 * attr(pca_data, "percentVar"))
  
  ggplot(pca_data, aes(PC1, PC2, color = group)) +
    geom_point(size = 3) +
    labs(
      x = paste0("PC1: ", pct[1], "% variance"),
      y = paste0("PC2: ", pct[2], "% variance"),
      color = NULL
    ) +
    theme_minimal()
}
```

### MCTi

TGFβ drives the primary axis of variation. Inhibitor effects are visible along
PC2 after batch correction.

```{r pca-mcti, fig.cap = "PCA of the MCTi experiment after batch correction."}
plot_pca(dds_mcti, remove_batch = TRUE)
```

### VB253

```{r pca-vb253, fig.cap = "PCA of the VB253 experiment."}
plot_pca(dds_vb253)
```

## Differential expression

We extract results for each comparison of interest using the group contrasts,
then apply `ashr` log-fold-change shrinkage. The unshrunken Wald statistic is
retained for GSEA ranking.

```{r de-function}
get_results <- function(dds, contrast, alpha = 0.05) {
  res <- results(dds, contrast = contrast, alpha = alpha)
  res_shrunk <- lfcShrink(dds, res = res, type = "ashr")
  
  as_tibble(res_shrunk, rownames = "row") |>
    mutate(stat = res$stat) |>
    left_join(
      as_tibble(rowData(dds)[, c("entrezid", "symbol")], rownames = "row"),
      by = "row"
    ) |>
    relocate(symbol, entrezid, .after = row) |>
    arrange(padj)
}
```

### MCTi contrasts

```{r de-mcti}
deg_tgfb <- get_results(dds_mcti, c("group", "TGFβ.VEH", "CTL.VEH"))
deg_azd  <- get_results(dds_mcti, c("group", "TGFβ.AZD", "TGFβ.VEH"))
deg_vb   <- get_results(dds_mcti, c("group", "TGFβ.VB", "TGFβ.VEH"))
deg_dual <- get_results(dds_mcti, c("group", "TGFβ.AZD.VB", "TGFβ.VEH"))
```

### VB253 contrasts

```{r de-vb253}
deg_tgfb_vb253 <- get_results(dds_vb253, c("group", "TGFβ.VEH", "CTL.VEH"))
deg_vb253      <- get_results(dds_vb253, c("group", "TGFβ.VB253", "TGFβ.VEH"))
```

### Summary

```{r de-summary}
tibble(
  comparison = c(
    "TGFβ vs CTL (MCTi)", "AZD vs VEH", "VB vs VEH",
    "AZD/VB vs VEH", "TGFβ vs CTL (VB253)", "VB253 vs VEH"
  ),
  significant = c(
    sum(deg_tgfb$padj < 0.05, na.rm = TRUE),
    sum(deg_azd$padj < 0.05, na.rm = TRUE),
    sum(deg_vb$padj < 0.05, na.rm = TRUE),
    sum(deg_dual$padj < 0.05, na.rm = TRUE),
    sum(deg_tgfb_vb253$padj < 0.05, na.rm = TRUE),
    sum(deg_vb253$padj < 0.05, na.rm = TRUE)
  )
)
```

## Volcano plots

```{r volcano-function}
plot_volcano <- function(deg, title = "") {
  df <-
    deg |>
    filter(!is.na(padj), !is.na(symbol)) |>
    mutate(sig = if_else(padj < 0.05, "sig", "ns"))
  
  top <-
    df |>
    filter(sig == "sig") |>
    slice_min(padj, n = 20)
  
  ggplot(df, aes(log2FoldChange, -log10(padj))) +
    geom_point(aes(color = sig), size = 0.5, alpha = 0.5) +
    ggrepel::geom_text_repel(
      data = top,
      aes(label = symbol),
      size = 2.5,
      max.overlaps = 15,
      segment.size = 0.2
    ) +
    scale_color_manual(values = c(ns = "grey70", sig = "darkred")) +
    labs(title = title, x = "log2 Fold Change", y = "-log10 Adjusted P") +
    theme_minimal() +
    theme(legend.position = "none")
}
```

```{r volcano-mcti, fig.height = 8, fig.width = 7, fig.cap = "Volcano plots for MCTi contrasts."}
cowplot::plot_grid(
  plot_volcano(deg_tgfb, "TGFβ vs CTL"),
  plot_volcano(deg_azd, "AZD vs VEH in TGFβ"),
  plot_volcano(deg_vb, "VB vs VEH in TGFβ"),
  plot_volcano(deg_dual, "AZD/VB vs VEH in TGFβ"),
  ncol = 2
)
```

```{r volcano-vb253, fig.cap = "Volcano plots for VB253 contrasts."}
cowplot::plot_grid(
  plot_volcano(deg_tgfb_vb253, "TGFβ vs CTL"),
  plot_volcano(deg_vb253, "VB253 vs VEH in TGFβ"),
  ncol = 2
)
```

## Gene set enrichment analysis

GSEA is performed using the `fgsea` package with Hallmark gene sets from
MSigDB. Genes are ranked by the Wald statistic from DESeq2.

```{r gsea-setup}
hallmark <-
  msigdbr::msigdbr(species = "Homo sapiens", category = "H") |>
  select(gs_name, ensembl_gene) |>
  group_by(gs_name) |>
  summarise(genes = list(ensembl_gene), .groups = "drop") |>
  deframe()
```

```{r gsea-function}
run_gsea <- function(deg, pathways) {
  ranks <-
    deg |>
    filter(!is.na(stat)) |>
    select(row, stat) |>
    arrange(stat) |>
    deframe()
  
  fgsea::fgsea(pathways = pathways, stats = ranks, eps = 0,
               nPermSimple = 10000) |>
    as_tibble() |>
    filter(padj < 0.05) |>
    mutate(pathway = str_remove(pathway, "^HALLMARK_")) |>
    arrange(desc(NES))
}
```

### MCTi

```{r gsea-mcti}
gsea_tgfb <- run_gsea(deg_tgfb, hallmark)
gsea_azd  <- run_gsea(deg_azd, hallmark)
gsea_vb   <- run_gsea(deg_vb, hallmark)
gsea_dual <- run_gsea(deg_dual, hallmark)
```

### VB253

```{r gsea-vb253}
gsea_tgfb_vb253 <- run_gsea(deg_tgfb_vb253, hallmark)
gsea_vb253      <- run_gsea(deg_vb253, hallmark)
```

### Hallmark pathway dot plot

This visualization compares pathway enrichment across the MCTi drug conditions.

```{r gsea-dotplot, fig.height = 8, fig.width = 6, fig.cap = "GSEA dot plot comparing Hallmark pathway enrichment across drug conditions."}
gsea_combined <-
  list(
    "TGFβ" = gsea_tgfb,
    "AZD" = gsea_azd,
    "VB" = gsea_vb,
    "AZD/VB" = gsea_dual
  ) |>
  bind_rows(.id = "comparison") |>
  mutate(
    pathway = str_replace_all(pathway, "_", " "),
    comparison = factor(comparison, levels = c("TGFβ", "AZD", "VB", "AZD/VB"))
  )

ord <-
  gsea_combined |>
  select(pathway, comparison, NES) |>
  pivot_wider(names_from = "comparison", values_from = "NES") |>
  arrange(
    desc("TGFβ"),
    desc("AZD"),
    desc("VB"),
    desc("AZD/VB")
  ) |>
  pull(pathway)

gsea_combined |> 
  mutate(pathway = factor(pathway, levels = ord)) |> 
  ggplot(aes(comparison, pathway)) +
  geom_point(aes(fill = NES), shape = 21, size = 3, color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    limits = c(-2, 2), oob = scales::squish
  ) +
  scale_y_discrete(limits = rev) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 7),
    panel.grid.major.y = element_line(linewidth = 0.25, color = "grey90")
  )
```

## Transcription factor enrichment analysis

TFEA uses decoupleR with DoRothEA regulons (confidence levels A and B) to
infer transcription factor activities from normalized counts. Activity scores
are then tested across experimental groups using limma.

```{r tfea-functions}
prep_count_matrix <- function(dds) {
  rd <-
    as_tibble(rowData(dds), rownames = "row") |>
    filter(!is.na(symbol), symbol != "") |>
    group_by(symbol) |>
    slice_max(baseMean, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(row, symbol)
  
  mat <- counts(dds, normalized = TRUE)
  mat <- mat[rd$row, , drop = FALSE]
  rownames(mat) <- rd$symbol
  mat
}

run_tfea <- function(dds) {
  mat <- prep_count_matrix(dds)
  
  regulons <-
    dorothea::dorothea_hs |>
    filter(confidence %in% c("A", "B")) |>
    rename(source = tf)
  
  decoupleR::run_viper(
    mat = mat,
    network = regulons,
    .source = "source",
    .target = "target",
    .mor = "mor",
    minsize = 5
  )
}

pivot_tfea_scores <- function(tfea_result) {
  tfea_result |>
    filter(statistic == "viper") |>
    select(source, condition, score) |>
    pivot_wider(names_from = condition, values_from = score) |>
    column_to_rownames("source") |>
    as.matrix()
}
```

```{r tfea-run}
tfea_mcti <- run_tfea(dds_mcti)
tfea_vb253 <- run_tfea(dds_vb253)
```

### Limma analysis of TF activity scores

TF activity scores are assembled into a matrix and tested with limma contrasts
to identify transcription factors with differential activity between conditions.
The MCTi experiment uses `duplicateCorrelation` to account for replicate
blocking.

```{r tfea-limma}
fit_tf_limma <- function(scores, dds, contrasts_list, block = NULL) {
  cd <- as_tibble(colData(dds), rownames = "sample_id")
  scores <- scores[, cd$sample_id, drop = FALSE]
  
  design <- model.matrix(~ 0 + group, data = colData(dds))
  colnames(design) <- str_remove(colnames(design), "^group")
  
  cm <- do.call(limma::makeContrasts, c(contrasts_list, list(levels = design)))
  
  if (!is.null(block)) {
    corfit <- limma::duplicateCorrelation(scores, design, block = block)
    fit <- limma::lmFit(scores, design,
                        block = block,
                        correlation = corfit$consensus)
  } else {
    fit <- limma::lmFit(scores, design)
  }
  
  fit |>
    limma::contrasts.fit(cm) |>
    limma::eBayes()
}
```

```{r tfea-mcti-fit}
scores_mcti <- pivot_tfea_scores(tfea_mcti)

fit_mcti <- fit_tf_limma(
  scores_mcti,
  dds_mcti,
  contrasts_list = list(
    tgfb = "TGFβ.VEH - CTL.VEH",
    azd  = "TGFβ.AZD - TGFβ.VEH",
    vb   = "TGFβ.VB - TGFβ.VEH",
    dual = "TGFβ.AZD.VB - TGFβ.VEH"
  ),
  block = dds_mcti$replicate
)

limma::topTable(fit_mcti, coef = "tgfb", number = 10) |>
  as_tibble(rownames = "tf")
```

```{r tfea-vb253-fit}
scores_vb253 <- pivot_tfea_scores(tfea_vb253)

fit_vb253 <- fit_tf_limma(
  scores_vb253,
  dds_vb253,
  contrasts_list = list(
    tgfb  = "TGFβ.VEH - CTL.VEH",
    vb253 = "TGFβ.VB253 - TGFβ.VEH"
  )
)

limma::topTable(fit_vb253, coef = "vb253", number = 10) |>
  as_tibble(rownames = "tf")
```

## Session info

```{r session-info, eval = TRUE}
sessionInfo()
```
